---
title: "Data Version Summary Stats"
output:
  html_document:
    toc: True
    toc_float: True
    df_print: paged
---

```{r setup, include=FALSE}
# packages
library(tidyverse)
library(knitr)

# knitr settings
opts_chunk$set(  options(scipen=999), rows.print=25 )
knit_hooks$set(inline = function(x) { prettyNum(x, big.mark=",") })

# directories
source("./R/0_utils.R")
```

```{r utils, include=FALSE}
# UTILITY FUNCTIONS

pass.to.global <- function(this.list){
  #' passes every element of a names list to the global environment

  for(item in 1:length(this.list)){ 
    assign(x = names(this.list)[item], value = this.list[[item]], envir = globalenv()) 
  }
  
}

prep.data <- function(){
  #' loads a PanelPRO and BayesMendel final data .RData objects,
  #' splits them into chunks by data type (demographics, surgeries, cancer history,
  #' marker tests, and genetic tests) and loads those chunks in the global environment.
  
  load(paste0(final.date.dir,"clean-COH-data-PanelPRO.RData"))
  load(paste0(final.date.dir,"clean-COH-data-BayesMendel.RData"))
  
  # show column names and row counts
  print(paste0("PanelPRO Rows: ", nrow(pp.clean)))
  print("PanelPRO Columns:")
  print(colnames(pp.clean))
  cat("\n")
  print(paste0("BayesMendel Rows: ", nrow(bm.clean)))
  print("BayesMendel Columns:")
  print(colnames(bm.clean))
  
  # demographics data
  demo.dat <- 
    pp.clean %>%
    select(SubjectID,
           PedigreeID,
           Region,
           ID,
           Sex,
           MotherID,
           FatherID,
           isProband,
           CurAge,
           isDead,
           race,
           Ancestry,
           isHispanic,
           Twins)
  
  bm.demo <- select(bm.clean, SubjectID, race, ethnic)
  colnames(bm.demo)[2:3] <- c("bm.race","bm.ethnic")
  demo.dat <- left_join(demo.dat, bm.demo, by = "SubjectID")
  
  # cancer history data
  cancer.dat <- 
    pp.clean %>%
    select(SubjectID,
           PedigreeID,
           starts_with("isAff"),
           starts_with("Age"))
  
  surg.dat <-
    pp.clean %>%
    select(SubjectID,
           PedigreeID,
           riskmod,
           interAge)
  
  bm.surg.ages <- select(bm.clean, SubjectID, 
                         Mastectomy, AgeMastectomy, 
                         Oophorectomy, AgeOophorectomy)
  surg.dat <- left_join(surg.dat, bm.surg.ages, by = "SubjectID")
  
  # marker test (and colorectal cancer location) data
  marker.dat <-
    pp.clean %>%
    select(SubjectID,
           PedigreeID,
           ER,
           CK14,
           CK5.6,
           PR,
           HER2,
           MSI)
  
  bm.mark <- select(bm.clean, SubjectID, location)
  marker.dat <- left_join(marker.dat, bm.mark, by = "SubjectID")
  
  # genetic test data
  test.dat <-
    pp.clean %>%
    select(SubjectID,
           PedigreeID,
           CarrierStatus,
           APC:TP53)
  
  p16 <- select(bm.clean, SubjectID, P16)
  
  # pass to global environment
  df.list <- list(demo.dat = demo.dat, 
                  cancer.dat = cancer.dat, 
                  surg.dat = surg.dat, 
                  marker.dat = marker.dat, 
                  test.dat = test.dat,
                  p16=p16)
  
  pass.to.global(df.list)
  
}

### Summary Functions

demo.summary <- function(df = demo.dat){
  #' calculate and return summary statistics for demographics data

  # tables
  region <- table(df$Region, useNA = "always")
  sex <- table(df$Sex, useNA = "always")
  races <- table(df$race, useNA = "always")
  bm.races <- table(df$bm.race, useNA = "always")
  ancestry <- table(df$Ancestry, useNA = "always")
  bm.ancestry <- table(df$bm.ethnic, useNA = "always")
  
  # counts
  peds <- length(unique(df$PedigreeID))
  subs <- length(unique(df$SubjectID))
  pros <- sum(df$isProband)
  rels <- subs - pros
  hisp <- sum(df$isHispanic, na.rm = T)
  deaths <- sum(df$isDead)
  twins <- length(df$Twins[which(df$Twins != 0)])
  
  # numeric summaries
  ages <- summary(df$CurAge)
  ped.size <- summary(df %>%
                      select(PedigreeID) %>%
                      group_by(PedigreeID) %>%
                      summarise(PedSize = n()) %>%
                      select(PedSize) %>%
                      as.matrix() %>%
                      as.vector())
  
  # missingness by variable
  missing.demo <- apply(df, 2, function(x) sum(is.na(x)))
  
  # pass to global environment
  ld.list <- list(subs=subs, peds=peds, ped.size=ped.size, region=region, 
                  sex=sex, pros=pros, rels=rels, ages=ages, 
                  deaths=deaths, races=races, bm.races=bm.races, 
                  ancestry=ancestry, bm.ancestry=bm.ancestry, hisp=hisp, 
                  twins=twins, missing.demo=missing.demo)
  
  pass.to.global(ld.list)
  
}

cancer.summary <- function(df = cancer.dat){
  #' calculate summary stats of cancer history and pass to global
  
  # cancer status (minus CBC & BC2)
  table.cancer <- apply(df %>% select(starts_with("isAff")), 2, function(x) table(x, useNA = "always"))
  df.cancer <- as.data.frame.list(table.cancer[-((length(table.cancer)-1):length(table.cancer))])
  colnames(df.cancer)[1] <- "AffectedStatus"
  df.cancer <- select(df.cancer, -ends_with(".x"))
  colnames(df.cancer)[2:length(colnames(df.cancer))] <- sub("isAff","",colnames(df.cancer)[2:length(colnames(df.cancer))])
  colnames(df.cancer)[2:length(colnames(df.cancer))] <- sub(".Freq","",colnames(df.cancer)[2:length(colnames(df.cancer))])
  df.cancer <- as.data.frame(t(df.cancer))
  colnames(df.cancer)[1:2] <- df.cancer[1,1:2]
  colnames(df.cancer)[3] <- "NA"
  df.cancer <- df.cancer[-1,]
  
  # Cancer ages
  cancer.age.summary <- round(apply(df %>% select(starts_with("Age")), 2, function(x) summary(x)),1)
  
  # BC2 status
  cbc.table <- table(df$isAffBC2, useNA = "always")
  cbc.bc.bc2.diff <- 
    df %>%
    select(AgeBC, AgeBC2) %>%
    mutate(YrsBetween = AgeBC2 - AgeBC) %>%
    mutate(ModYrsBetween = ifelse(YrsBetween > 10, "11+", YrsBetween))
  cbc.age.diff.summary <- round(summary(cbc.bc.bc2.diff$YrsBetween), 1)
  cbc.age.diff.table <- table(cbc.bc.bc2.diff$ModYrsBetween)
  cbc.age.diff.table <- cbc.age.diff.table[c(1:2,5:12,3:4)]
  
  
  # pass to global
  ld.list <- list(df.cancer=df.cancer,
                  cancer.age.summary=cancer.age.summary,
                  cbc.table=cbc.table,
                  cbc.age.diff.summary=cbc.age.diff.summary,
                  cbc.age.diff.table=cbc.age.diff.table)
  
  pass.to.global(ld.list)
}

surg.summary <- function(df = surg.dat){
  #' calculate summary stats for prophylactic surgeries and pass to global
  
  surgs <- df %>% filter(!is.na(riskmod))
  
  # number of subjects with at least one surgery
  surg.subs.cnt <- nrow(surgs)
  
  # number of surgeries
  mast.cnt <- sum(grepl(pattern = "Mastectomy", surgs$riskmod))
  ooph.cnt <- sum(grepl(pattern = "Oophorectomy", surgs$riskmod))
  
  # summaries of surgery ages
  mast.ages <- summary(df %>%
                        select(Mastectomy,AgeMastectomy) %>%
                        filter(Mastectomy == 1) %>%
                        select(AgeMastectomy) %>%
                        as.matrix() %>%
                        as.vector())
  
  ooph.ages <- summary(df %>%
                        select(Oophorectomy,AgeOophorectomy) %>%
                        filter(Oophorectomy == 1) %>%
                        select(AgeOophorectomy) %>%
                        as.matrix() %>%
                        as.vector())
  
  # pass to global
  ld.list <- list(surg.subs.cnt=surg.subs.cnt,
                  mast.cnt=mast.cnt,
                  ooph.cnt=ooph.cnt,
                  mast.ages=mast.ages,
                  ooph.ages=ooph.ages)
  
  pass.to.global(ld.list)
  
}

marker.summary <- function(df = marker.dat){
  #' calculate bio marker summary stats and pass to global environment
  
  # result type counts by bio marker
  marker.table <- apply(df %>% select(-c(SubjectID,PedigreeID,location)), 2, function(x) table(x, useNA = "always"))
  
  # table of colorectal cancer locations (for BayesMendel)
  loc.table <- table(df$location, useNA = "always")
  
  # pass to global
  ld.list <- list(marker.table=marker.table,
                  loc.table=loc.table)
  
  pass.to.global(ld.list)
  
}

test.summary <- function(df = test.dat, bm.p16 = p16){
  #' calculate gene test summary stats and pass to global
  
  not.tested   <- sum(is.na(test.dat$CarrierStatus))
  pos.carriers <- sum(test.dat$CarrierStatus[which(!is.na(test.dat$CarrierStatus))] == 1)
  vus.carriers <- sum(test.dat$CarrierStatus[which(!is.na(test.dat$CarrierStatus))] == 2)
  non.carriers <- sum(test.dat$CarrierStatus[which(!is.na(test.dat$CarrierStatus))] == 0)
  
  tested.per.ped <- summary(test.dat %>%
                            select(PedigreeID,CarrierStatus) %>%
                            filter(!is.na(CarrierStatus)) %>%
                            group_by(PedigreeID) %>%
                            summarise(Count = n()) %>%
                            select(Count) %>%
                            as.matrix() %>%
                            as.vector())
  
  gene.results <- apply(test.dat %>% select(-c(SubjectID,PedigreeID,CarrierStatus)), 
                        2, 
                        function(x) table(x, useNA = "always"))
  
  p16.results <- table(bm.p16$P16, useNA = "always")
  
  
  # pass to global
  ld.list <- list(not.tested=not.tested,
                  pos.carriers=pos.carriers,
                  vus.carriers=vus.carriers,
                  non.carriers=non.carriers,
                  tested.per.ped=tested.per.ped,
                  gene.results=gene.results,
                  p16.results=p16.results)
  
  pass.to.global(ld.list)
  
}

all.gene.summary <- function(){
  #' table of results by gene type for all genes in COH data set
  
  # load all genes data set
  load(paste0(final.date.dir,"clean-COH-data-all-genes.RData"))
  gene.results <- dat.clean %>% select(ABL1:ZRSR2)
  
  # counts by result type and gene
  pos.cnts <- apply(gene.results, 2, function(x) sum(!is.na(x) & x==1, na.rm = T))
  vus.cnts <- apply(gene.results, 2, function(x) sum(!is.na(x) & x==2, na.rm = T))
  neg.cnts <- apply(gene.results, 2, function(x) sum(!is.na(x) & x==0, na.rm = T))
  no.test.cnts <- apply(gene.results, 2, function(x) sum(is.na(x)))
  
  # put it in a table
  cnts <- as.data.frame(rbind(pos.cnts, vus.cnts, neg.cnts, no.test.cnts))
  rownames(cnts) <- c("Pos", "VUS", "Neg", "No Test")
  cnts
  
}
```


# Data

**Date of Data Generation / Version: `r todays.date`**

```{r data_source, echo=FALSE}
# load and prepare data PanelPRO and BayesMendel data frames
prep.data()
```


# Demographics

```{r demo, echo=FALSE}
demo.summary()
```

Subjects: `r subs`

Pedigrees: `r peds`

Pedigree Size:
```{r ped_size, echo=FALSE}
ped.size
```

Region:
```{r region, echo=FALSE}
region
```

Sex:
```{r sex, echo=FALSE}
sex
```

Probands: `r pros`

Ages:
```{r ages, echo=FALSE}
ages
```

Deaths: `r deaths`

PanelPRO Races:
```{r races, echo=FALSE}
races
```

BayesMendel Races (no missing values due to single race category for each family requirement):
```{r bm_races, echo=FALSE}
bm.races
```

PanelPRO Ancestry:
```{r ancestry, echo=FALSE}
ancestry
```

BayesMendel ethnic:
```{r bm_ancestry, echo=FALSE}
bm.ancestry
```

Hispanics: `r hisp`

Twins: `r twins`

Missingness:
```{r demo_missing, echo=FALSE}
missing.demo
```


# Cancer History

```{r cancer, echo=FALSE}
cancer.summary()
```

**Cancer Status (excluding CBC/BC2)** (0 = no cancer, 1 = cancer)
```{r cancer_status, echo=FALSE}
df.cancer
```

Cancer Ages:
```{r cancer_ages, echo=FALSE}
cancer.age.summary
```


2nd Breast Cancer Status:
```{r cbc, echo=FALSE}
cbc.table
```

Summary of Difference in Years between 1st BC and 2nd BC:
```{r cbc-age-diff, echo=FALSE}
cbc.age.diff.summary
```

Table of Difference in Years between 1st BC and 2nd BC:
```{r cbc-age-tbl, echo=FALSE}
cbc.age.diff.table
```

# Prophylactic Surgeries

```{r surg, echo=FALSE}
surg.summary()
```

Subjects with a prophylactic surgery: `r surg.subs.cnt`

Mastectomies: `r mast.cnt`

Mastectomy Ages:
```{r mast_ages, echo=FALSE}
mast.ages
```

Oophorectomies: `r ooph.cnt`

Oophorectomy Ages:
```{r ooph_ages, echo=FALSE}
ooph.ages
```

# Marker Tests

```{r marker, echo=FALSE}
marker.summary()
```

Marker Results: 
```{r marker_results, echo=FALSE}
marker.table
```

Colorectal Cancer Location:
```{r location, echo=FALSE}
loc.table
```


# Gene Tests

```{r test, echo=FALSE}
test.summary()
```

Subjects with at least 1 positive gene: `r pos.carriers`

Subjects with at least 1 VUS gene and no positive genes: `r vus.carriers`

Subjects with all negative tests: `r non.carriers`

Subjects never tested: `r not.tested`

Subjects with Gene Tests Per Pedigree:
```{r test_by_ped, echo=FALSE}
tested.per.ped
```

Results by PanelPRO Gene (NA = no test, 0 = neg, 1 = pos, 2 = VUS):
```{r gene_results, echo=FALSE}
gene.results
```

P16 Results:
```{r p16, echo=FALSE}
p16.results
```

**All Gene Results in Study**
```{r all_genes, echo=FALSE}
all.genes <- all.gene.summary()
pander::pander(all.genes, split.table = 80, style = "rmarkdown")
```





